A Mechanised Library for Generic (Co)Programming

### 2023-12-31 paper/paper.tex
some edits to the intro, related work with more citations.

\maketitle
\section{Introduction}
Recursive definitions cannot be proven well-defined automatically. Modern proof
assistants like Coq or Agda provide a sound, but incomplete algorithm which
automatically checks for termination or productivity. For recursion, this is
done by trying to (automatically) infer which argument in the recursive call
gets smaller with respects to the original input argument. For productivity, the
algorithm checks that the (co)recursive call appears directly under a
constructor to make sure that this function always produce at least one element
after each recursive step.

As mentioned, these algorithms are incomplete meaning that some functions,
though well-defined, cannot be recognized so by the proof assistant.

One example of this fact is the quicksort algorithm given by the
\haskell{partition} and \haskell{quicksort} function below (written in Haskell
code):
\begin{minted}{haskell}
partition :: [a] -> ([a], a, [a])
partition (x:xs) = ((filter (x <=) xs), x, (filter (x >) xs))

quicksort :: [a] -> [a]
quicksort [] = []
quicksort ls = let (xs, x, ys) = partition ls in
                        (quicksort xs) ++ [x] ++ (quicksort ys)
\end{minted}
While \haskell{quicksort} is a well-defined mathematical function, a proof
assistant such as Agda or Coq is not able to accept such definitions. The reason
is that the function \emph{destructuring} the input, that is the
\haskell{partition} function is not simply destructuring the list returning the
head and the tail of it, but it returns two sublists with the head as a pivot.

While the syntactic checker of a proof assistant cannot understand that these
two sublists are eventually going to be smaller than the original input we can
can give a pen-and-paper by showing the \haskell{partition} is a
\emph{well-founded coalgebra} on the functor $(- \times A \times -)$ for a fixed
type $A$. Here the term ``well-founded'' refers to the fact that the infinite
iteration of this coalgebra will produce a \emph{finite} binary tree with nodes
in $A$.

@@

With this work we intend to begin a study of recursion scheme beyond the world
of functional programming to the one of dependent type theories such as a the
ones employed in proof assistants like Agda or Coq where non-termination is not
included in the language. For type theories it has been a long-standing
challenge to come up with a sound algorithm which rejects all programs that do
not terminate (soundness) and accepts as many well-defined programs as possible
(completeness). Our aim is to provide an additional tool to the every-day
programmer to be able to extend the range of recursive programs we can write.

@@

>>>>

\maketitle
\section{Introduction}
Recursive definitions cannot be proven well-defined automatically due to the
halting problem. Modern proof assistants like Coq or Agda provide a sound, but
incomplete algorithm which syntactically checks for termination or productivity.
For recursion, this is done by trying to (automatically) infer which argument in
the recursive call gets smaller with respects to the original input argument.
For productivity, the algorithm checks that the (co)recursive call appears
directly under a constructor to make sure that this function always produce at
least one element after each recursive step.

As mentioned, some functions, though well-defined, cannot be accepted by the
proof assistant. One example of this fact is the quicksort algorithm given by
the \haskell{partition} and \haskell{quicksort} function below (written in
Haskell code):
\begin{minted}{haskell}
partition :: [a] -> ([a], a, [a])
partition (x:xs) = (filter (x <=) xs, x, filter (x >) xs)

quicksort :: [a] -> [a]
quicksort [] = []
quicksort ls = let (xs, x, ys) = partition ls in
                        (quicksort xs) ++ [x] ++ (quicksort ys)
\end{minted}
While \haskell{quicksort} is a well-defined mathematical function it cannot be
accepted by a proof assistant.  The reason is that the \haskell{partition}
function \emph{destructuring} the input ``dives'' deeper in the input returning
two sublists with the head as a pivot.

A pen-and-paper technique would be to show that the \haskell{partition} is a
\emph{well-founded coalgebra} on the functor $(- \times A \times -)$ for a fixed
type $A$. Here the term ``well-founded'' refers to the fact that the infinite
iteration of this coalgebra will produce a \emph{finite} binary tree with nodes
in $A$.

@@

With this work we intend to begin a study of the applications of recursion
schemes that goes beyond functional programming, that is to dependent type
theories such as a the ones employed in proof assistants like Agda or Coq where
non-termination is outright rejected.  This has been a long-standing challenge
to come up with a sound algorithm which rejects all programs that do not
terminate (soundness) and accepts as many well-defined programs as possible
(completeness). Our aim is to provide an additional tool to the every-day
programmer to be able to extend the range of recursive programs we can write.

@@

\section{Related Work}

\begin{itemize}
  \item McBride devised a generalisation of Bove and Capretta's general
recursive function monad~\cite{McBride15, BoveC01}
  \item Guarded recursion for productive definitions and general recursion
~\cite{AtkeyM13, PaviottiMB15}
  \item PACO library~\cite{HurNDV13}
  \item Sized Types \cite{AbelPTS13, AbelP16}
\end{itemize}
